### 2019-03-31 response/response-demography-1.tex

I have rewritten the Introduction entirely to situate the paper in the literature 
and better highlight the contributions of the paper.
I now cover the research on spacing as a measure of son preference, the 
relationship between spacing and child health (both older and younger child), the 
effect on women's health and other outcomes, and, finally, the limited and mixed results
on what determines birth spacing behavior in developing countries.
I use these to motivate why it is essential to examine how birth spacing has been affected 
by the use of sex selection, and, in addition, highlight the three main questions I address 
in the paper: the effect of sex selection on birth intervals, the distribution of births 
within spells, and, finally, differences across geographical regions.

\item Second, a conceptual framework or theoretical argument is needed to
motivate the decision to examine birth intervals and sex composition by
maternal education. Does maternal education proxy for access to sex
selective technology, preferences for sons, etc.? 
In addition, how does
selection into the highest/lowest educated categories change over the 45
years of the study? How could this impact the interpretation of results?

[Response:]

I have expanded the discussion of the choice of non-proportionality variables in the 
``Explanatory Variables'' section (and added a footnote to that effect in the ``Estimation 
Strategy'' section).
The discussion now covers education, sex composition, and area of residence in more detail,
although given that I need to keep the paper below 8,000 words there is a limit to how
detailed the discussion is.
If there is a more appropriate place for this discussion, I am happy to move it.

@@

When it comes to selection, I see two possible explanations: one economic and one 
cultural.
In the ``economic'' explanation, well-educated women's use of sex selection is mainly 
driven by the lower fertility that comes from their higher opportunity cost of time and 
the higher income that allow them to use sex selection.
In that case, there is no selection bias since the use of sex selection is driven by 
factors that are identical across those with the same level of education.

My ``cultural'' explanation assumes that, originally, those with higher education were also 
those groups with the strongest son preference (presumably also higher caste).
The expansion of access to education brought other groups/castes into connection with
the higher castes and led them to assimilate the high caste son preference norms.
Presumably, this assimilation would not be complete, and those groups newer to higher 
education would, therefore, have lower son preference and therefore lower use of sex 
selection.
Hence, any appearance of lower use of sex selection could be a selection story,
and that the initial group's son preference is just as strong.
The cultural explanation also suggests a reason for why the stated son preference appears 
to decline over time: the ``newer'' entrants have lower initial son preference and, 
therefore, the average son preference for high education women will decline over time
\citep{bhat03,pande07}.

Since I have no way of separating these two possible explanations with the available
data and set up and because of the need to keep the paper short, I have not included this 
discussion in the paper itself. 
One possible approach would be to include this as a question for future research in the
Conclusion.
If you think it will make the paper stronger, I am, of course, open to including the
discussion there or anywhere else it might fit better.

@@

Both NFHS-3 and NFHS-4 ask for the specific month and year of first marriage, but some
respondents did not provide complete information.
One approach to understanding the impact of imputation on spacing to first birth is to 
split spacing from marriage to first birth by completeness of date of marriage.
In NFHS-3, 2.55 percent of the 39,213 women with complete information on the date of 
marriage shows a spacing from marriage to first birth of 8 months or less and that 0.6 
percent show a negative spacing. 
For the 19,403 women with incomplete information on the date of marriage, 27.36 percent show
a spacing from marriage to first birth of eight months or fewer and that 3.37 percent show a 
negative duration.
In NFHS-4, 3.8 percent of women with complete information report a birth eight or fewer months
after marriage and 2.35 report a negative duration.
For women with imputed information, 14.99 percent report a birth eight or fewer months
after marriage with 11.02 percent reporting a negative duration.
The main difference between NFHS-3 and NFHS-4 is that only just below four percent of 
women in NFHS-4 have imputed information whereas more than 33 percent did in NFHS-3.

Given the substantial problems with the first spell data for all but the very last
survey and that the first spell analysis adds relatively little to the central question, 
I have dropped the analysis of first spells.
Removing the first spell also helps to keep the paper within the word limit.
Because of the large number of imputed observations for the first spell, I have also
remove the first spell duration and its square as explanatory variables for the second
through fourth spells.

@@

I have added a detailed discussion of recall error in the Appendices intended for online 
publication.
Furthermore, I have changed the cut-off to be 22 years of marriage or more for all surveys 
to simplify.
Even with the stricter cut-off, there is, however, still some evidence of recall error 
where the sex ratio was substantially biased against females  before sex selection became 
available.

\item I understand the empirical reason for not increasing the number of
variables. However, a key variable that is essential to consider is
region. This is important because regional influences on birth spacing
and sex selection are very strong as established in the literature. Sex
selection and abortion is especially strong in certain regions but not
in others. The regional differentials in fertility transition,
contraceptive use, spacing, sex ratios, abortions are clearly documented
in the literature. So it is vital to include this.

[Response:]

I agree that the regional differences in fertility, son preference, and use 
of sex selection are essential to examine.
To try to capture these differences, I have divided India into four broad regions 
depending on the degree of son preference and initial fertility level as in 
\citet{retherford03b}, although expanded to include all states in the surveys.
In addition to the power problem, some education/region/spell/period combinations have too 
small sample sizes for the estimations to converge.
The lack of convergence is primarily an issue for the lower and middle-education groups.
I, therefore, first present the all-India results and then focus on regional differences
for only the most educated women, who are the main users of sex selection.
This new section substitutes for the removed section on how sex ratios change within 
a spell.

### 2019-04-01 paper/sex_selection_spacing-ver2-demography.tex
Shorten version for Demography

\author{}
% \author{Claus C P\"ortner\\
%     Department of Economics\\
%     Albers School of Business and Economics\\
%     Seattle University, P.O. Box 222000\\
%     Seattle, WA 98122\\
%     \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
%     \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
%     \& \\
%     Center for Studies in Demography and Ecology \\
%     University of Washington\\ \vspace{2cm}
%     }

@@

Birth spacing has long served as a measure of son preference, with strong
son preference typically associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination has the potential to
change the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion automatically increases the interval between births by six 
months to a year.%
\footnote{
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Longer birth intervals in the absence of sons, which previously would be taken to indicate 
a lower son preference, now may arise precisely because a sufficiently strong son 
preference makes couples use sex selection.

In this paper, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce and apply an empirical method that directly incorporates the effects of 
sex-selective abortions on the duration between births 
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case since it has a long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
A significant proportion of Indian families exhibit differential stopping behavior, where, 
for a given number of children, they are more likely to have an additional child if they 
have not yet reached their preferred number of sons 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Moreover, mortality risk is higher for females than males for most ages groups, and
there has been an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.

% [What do we know about use of sex selection?]
Over the last three decades, India has seen dramatic increases in the males-to-females 
ratio at birth as access to prenatal sex determination expanded 
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2012,Portner2015b}.
Simultaneously, fertility has declined substantially, to the point where it is now at, or 
even below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
This decline in fertility, combined with a continued strong preference for sons, has 
intensified the pressure to use sex selection 
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In some instances, however, the fertility decline appears to have created a stronger 
aversion to daughters rather than a stronger preference for son \citep{DiamondSmith2008}.
}
The strongest predictor of the use of sex selection is the sex composition of previous 
children; for families without a son, the higher the parity, the higher is the probability
of having a son as the next birth \citep{retherford03b,jha06,abrevaya09,Portner2015b}.
The use of sex selection also increases with socioeconomic status, especially 
education, and is more prevalent in cities than in rural areas, both possibly because 
of lower desired fertility \citep{retherford03b,jha06,Portner2015b}.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still useful as a measure of son 
preference.
In India, the sex of the last-born child has in the past significantly affected the 
duration to next birth, with the expected birth interval about three percent longer if the 
last-born was a son than if it was a daughter \citep{Bhalotra2008,Kumar2016}.
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.%
\footnote{
Ethnic Indians in Malaysia do, however, show little evidence of son preference, although
the sample sizes are small \citep{Pong1994}.
}
In both Bangladesh and Vietnam, the more boys a family has the longer is the expected 
duration to the next birth \citep{Haughton1995,Haughton1996,Rahman1993,Soest2018}.
Outside Asia, the evidence is more mixed.
North Africa shows shorter spacing in the absence of sons, while a similar pattern does
not exist in Sub-Saharan Africa \citep{Rossi2015}.%
\footnote{
A possible exception is Senegal, where women with a higher risk of widowhood show
shorter birth spacing until they have secured a son \citep{Lambert2016}.
}

Second, birth spacing can have a substantial impact on children's outcomes.
Most of the research finds a negative impact of short spacing on child health---especially 
for very short intervals of 18 months or less---although identifying the causal mechanisms 
have proven more difficult \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There are three pathways through which birth intervals can affect health.
First, there is maternal depletion, where the mother does not have sufficient time to 
recover after her last pregnancy before she becomes pregnant again.%
\footnote{
This depletion can take the form of nutrition, folate, or cervical insufficiencies.
}
Second, disease transmission either from mother to child or between children.
Third, sibling competition, where siblings close in age compete for scarce resources and
parental care.
Which mechanism dominates is of particular interest when son preference drives spacing 
decisions.
Both the maternal depletion and the transmission of disease explanations predominately
affect the subsequent child and the mother.
Only sibling competition has the potential to affect the older child negatively.

In India, there is evidence that both the prior and the subsequent child have worse health 
outcomes and a higher likelihood of dying the shorter the interval between them, although 
the effects are not symmetrical
\citep{Whitworth2002,Bhargava2003,Rutstein2005,Maitra2008,Makepeace2008,Ghosh2018}.%
\footnote{
\cite{Bhalotra2008}, however, find that birth spacing only explains a limited amount
of neonatal mortality.
}
The increased mortality risk for prior children may come from shorter breastfeeding of 
girls to ensure that the mother can conceive again sooner, and with poor water
quality this shorter breastfeeding may result both in higher risk of dying and shorter
stature if the child survives \citep{Jayachandran2011,Jayachandran2017a}.
In other countries, focus has been mostly on the subsequent birth, but there is evidence 
in countries as diverse as Bangladesh, Brazil, El Salvadore, and Pakistan that
short intervals between births have adverse health effects
\citep{Cleland1984,Curtis1993,Davanzo2008,Gribble2009,Saha2013}.

The potential effects of spacing are not restricted to early outcomes, although
there is less evidence on longer-term effects and most comes from Western countries.
In the US, for example, a longer interval between births increases the older sibling's
test scores on the Peabody Individual Achievement Test, although there is no
statistically significant effect on the younger sibling \citep{Buckles2012}.
In line with this, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.
These results are, however, not uniformly supported, with recent results from Sweden 
showing no relationship between birth spacing, even if very short, and outcomes such as 
years of education completed, earnings, and unemployment \citep{Barclay2017}.

Third, the mother may also be affected by the length of the birth intervals.
The effect of spacing on mothers have received less attention than the effect on children, 
and reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are both
more likely to have short birth spacing and more likely to have anemia, possibly as
the result of the short spacing resulting in maternal depletion \citep{Milazzo2018}.
In Western countries, having a longer interval between births appear to have positive
effects on long-term labor market outcomes, such as participation and 
income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force in developing countries, understanding how couples make timing 
decisions will be necessary for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
For example, the evidence on the effects of education and socio-economic status on
birth intervals is mixed, with some research showing either a positive association, no
association, or one that is changing over time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.

Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births  \citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level of contraception, which 
results in longer spacing, to avoid having too many children by accident.
However, as contraception becomes more effective parents can more easily avoid future 
births, allowing them to reduce the spacing between births without having to worry about 
overshooting their preferred number of children.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult \citep{Yeakey2009}.%
\footnote{
There is evidence in Matlab that the effect of son preference on birth spacing is
stronger in areas with better access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time birth, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son 
preference in the literature, there has so far been no examination of what effect access 
to sex selection has had on spacing.
The overall hypothesis is that spacing patterns significantly changed as prenatal sex 
determination became available.
Specifically, I address three questions.

First, how did average birth intervals change over time for different groups and how are 
those changes related to the likely use of sex selection?
The underlying hypothesis is that in the absence of sex selection son preference leads to 
shorter spacing the fewer boys a family has for a given parity, whereas son preference 
increases the spacing the fewer boys a family has when prenatal sex selection is available.
I still expect parents to \emph{try} to conceive earlier after the birth of a girl than 
after a boy, but the abortions will increase the duration between births, and may even 
lead to longer average spacing after the birth of girls than after boys.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection may continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Very short birth intervals are, as mentioned, associated with worse health outcomes and 
increased mortality risk \citep{Conde-Agudelo2012}.
Hence, understanding how the distribution of birth intervals changes can help us understand to 
what extent we can expect improvements in health outcomes.
We should, for example, expect different impacts on the distribution of birth intervals
depending on whether parents change their spacing behavior based on an increasing 
concern for child health or from a strong son preference leading to continued pregnancies
until they conceive a son.

Finally, how are the changes in birth intervals with increasing access to sex selection 
distributed geographically?
Given the size and cultural differences within India, it is unlikely that the increasing 
access to sex selection will have had the same effects throughout \citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four regions.

Direct information on the use of sex selection is not available, but it is possible to 
compare different periods as a proxy since there have been substantial changes in access 
and legality of prenatal sex determination in India.
Abortion has been legal in India since 1971 and still is.
The first reports of sex determination appeared around 1982--83 
\citep{Sudha1999,bhat06,Grover2006}.
The number of clinics quickly increased, and knowledge about sex selection became 
widespread after a senior government official's wife aborted a fetus that turned out to be 
male \citep[p.\ 598]{Sudha1999}.
In 1994, the Central Government passed the Prenatal Diagnostic Techniques (PNDT) Act, 
making determining and communicating the sex of a fetus illegal.%
\footnote{
Details about the act are at \href{http://pndt.gov.in/}{http://pndt.gov.in/}.
The number of convictions has been low.
It took until January 2008 for the first state, Haryana, to reach five convictions.
Hence, private clinics operate with little risk of legal action \citep{Sudha1999}.
Furthermore, there is little evidence that bans like this significantly
affect sex ratios \citep{Das-Gupta2016}.
}

Based on these developments, I split the four decades into four periods.
The first period covers until prenatal sex determination became widely available, 
1972--1984.
The second period saw (mostly) unrestricted access to sex selection, 1985--1994.
Although, as reviewed above, there is evidence that there has been a substantial 
increase in the use of sex selection, some researchers argue that we might be close to or 
have already passed a turning-point in the use of sex selection 
\citep{Das_Gupta2009,Diamond-Smith2015}.
I, therefore, split the period from the implementation of the PNDT Act until the last 
survey into two, 1995--2004, and 2005--2016.

There are four main results.
First, there has been a general increase in the intervals between births for all education 
groups over the four decades covered by the data.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection, well-educated women with no
sons, have seen the largest increases in birth intervals.
As a result, we now have a complete reversal of the traditional spacing pattern for
well-educated women, and instead, observe the longest birth intervals for women with no 
sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
In other words, these women adhere to a strategy where they achieve a son through higher 
fertility rather than the use of sex selection.
Fourth, the traditional pattern of longer average spacing when women have one or more sons
appear to arise from more very long intervals, rather than from fewer very short birth 
intervals; across education groups there appear to be little difference in the likelihood
of having a very short birth interval across sex compositions, especially for lower
parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing to the birth of a girl or a boy.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth 
and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

I divide the explanatory variables into two groups.
The first group consists of variables expected to affect the shape of the hazard function
(the $\mathbf{Z}$ variables): 
mother's education, sex composition of previous children, and area of residence.
I chose these variables because the prior literature shows that they affect 
spacing choices and the likelihood of using sex selection.%
\footnote{
Increasing the number of variables interacted with the baseline hazard would
further lower the risk of bias but at the cost of requiring more data to 
precisely estimate.
}

The first variable that is likely to affect the hazard rate in a non-proportional manner
is education.
First, there is evidence, even in the absence of sex selection, that women with
different education levels have different hazard profiles, although the size and direction
of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Second, as discussed above, the use of sex selection increases with socioeconomic 
status, especially education, and the use of sex selection is unlikely to be the
same throughout a spell, which means that education cannot enter proportionally 
\citep{retherford03b,jha06,Portner2015b}.
One potential reason for the correlation between education and sex selection is that
increasing education of mothers correlates closely with lower fertility 
\citep{schultz97}.
This lower fertility leads, in turn, to a higher likelihood of using sex selection
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In principle, it is possible that part of the higher likelihood is driven by a stronger
son preference, independently of desired fertility.
There is, however, evidence that higher education women have shown a decline in
son preference \citep{bhat03,pande07}.
}
Another reason is that having more education is closely associated with higher income and 
wealth, which makes it easier to access illegal services such as prenatal sex 
determination.
It is not possible to separate these two effects here.

I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce
the potential problem of including other variables as proportional.%
\footnote{
Fathers' education has two opposite predicted effects: the associated higher income
should increase fertility and therefore lower the pressure to use sex selection, but
the higher income also makes the use of sex selection cheaper.
In practice, fathers' education had little effect on the hazards and the use of 
sex-selective abortions, and I, therefore, do not include it.
}

As discussed, the sex composition of previous children affects both the timing
of births and the use of sex-selective abortions 
\citep{retherford03b,jha06,Bhalotra2008,abrevaya09,Portner2015b,Kumar2016,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
As an example, for the third spell, three groups are used: Two boys,
one girl and one boy, and two girls.
It is, in principle, possible to estimate the model taking into account
the ordering of children, but this would further lower the power of the
test, by adding one additional group for the third spell and four
additional groups for the fourth spell.

>>>>

\author{}

% \author{Claus C P\"ortner\\
%     Department of Economics\\
%     Albers School of Business and Economics\\
%     Seattle University, P.O. Box 222000\\
%     Seattle, WA 98122\\
%     \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
%     \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
%     \& \\
%     Center for Studies in Demography and Ecology \\
%     University of Washington\\ \vspace{2cm}
%     }

@@

Birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals the
fewer sons a family has \citep{ben-porath76b,Leung1988}.
However, the introduction of prenatal sex determination potentially 
changed the relationship between son preference and birth spacing.
The stronger the son preference, the more likely couples are to resort to sex-selective 
abortions, and each abortion increases the interval between births by six to twelve 
months.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Hence, longer birth intervals in the absence of sons, which previously would be taken as 
an indicator of lower son preference, may now instead arise because couples with a strong 
son preference use sex selection.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I introduce an empirical method that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son. 
The method can be used to analyze both situations with and without prenatal
sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Why India?]
India is a particularly compelling case with its long history of male preference, 
especially in the northern states \citep{Kishor1993,murthi95,arnold98}.
Mortality risk is higher for females than males for most ages groups, resulting in
an almost continuous increase in India's overall ratio of males to females 
over the last century \citep{dyson01,Navaneetham2011,Bongaarts2015}.
Differential stopping behavior is widespread, with an additional child more likely the
fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.%
\footnote{
Other countries show a similar pattern
\citep[see, for example,][]{larsen98,filmer09,Altindag2016}.
}
Furthermore, fertility has declined substantially, to the point where it is now at, or even 
below, replacement in some areas 
\citep{Guilmoto2013,Dharmalingam2014,International-Institute-for-Population-Sciences-IIPS2017}.
The combination of falling fertility, strong son preference, and expanding access to 
prenatal sex determination has lead to dramatic increases in the males-to-females ratio at 
birth over the last three decades
\citep{das_gupta97,Sudha1999,Arnold2002,retherford03b,jha06,Guilmoto2009a,Guilmoto2012,Bongaarts2013,Portner2015b,Jayachandran2017}.

% [Why spacing?] 
There are four main reasons for examining birth spacing and how it changes with 
access to sex selection.
First, researchers have made extensive use of birth spacing as a measure of son preference, 
and it is critical to understand to what extent spacing is still a useful measure of son 
preference.
For example, in Bangladesh, India, and Vietnam, having more boys, and particularly having 
a boy as the last-born child, significantly increase the duration to next birth,
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
Ethnic Indians in South Africa also show a longer duration after the birth of a 
son than after a daughter \citep{Gangadharan2003}.
}
Outside Asia, the evidence is more mixed, with North Africa showing shorter spacing in the 
absence of sons, while a similar pattern does not exist in Sub-Saharan Africa 
\citep{Rossi2015}.

Second, birth spacing can affect children's short and long-term outcomes.
Most of the research finds a negative impact of short spacing on child health, especially 
for very short intervals of 18 months or less \citep{Conde-Agudelo2006,Conde-Agudelo2012}.
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
Most of the evidence on longer-term outcomes come from Western countries.
In the US, for example, a longer interval between births increases the older, but not the
younger, sibling's test scores \citep{Buckles2012}.
Furthermore, close spacing appears to increase the risk of dropping out of
high school and lower the probability of attending a post-secondary school in both the
US and Sweden \citep{Powell1993,Pettersson-Lidbom2009}.%
\footnote{
However, recent results from Sweden show no relationship between birth spacing, even if 
very short, and outcomes such as years of education completed, earnings, and unemployment 
\citep{Barclay2017}.
}

Third, the mother may also be affected by the length of the birth intervals, although 
reviews of the literature show mixed results on mother's anthropometric status 
\citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.
In Western countries, a longer interval between births has positive effects on long-term 
labor market outcomes, such as participation and income \citep{Gough2017,Karimi2014}.

Finally, we know less about what determines spacing behavior in developing countries than 
in developed countries, and with declining fertility and increasing numbers of women 
entering the labor force, understanding how couples make timing decisions is necessary 
for the design of suitable policies \citep{Portner2018}.%
\footnote{
It is clear that birth spacing does respond to policies in both developed and 
developing countries \citep{Pettersson-Lidbom2009,Todd2012,Meckel2015,Ghosh2018}.
}
Even the effects of access to contraceptive and declining fertility on spacing are unclear.
On the one hand, access to contraceptives allows women to avoid too short spacing between 
birth.
On the other hand, increased reliability of access and effectiveness of contraceptives can 
lead to shorter intervals between births if women used to have longer spacing to avoid
having too many children by accident \citep{Keyfitz1971,Heckman1976}.
These two counteracting effects may explain why better-educated women have shorter spacing 
than less educated women in some instances but not in others and why finding statistically 
significant effects of contraception use on birth spacing is difficult 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.%
\footnote{
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
Other evidence, however, points to families ability to time births, even in the absence
of modern contraceptives \citep{Jayachandran2011,Alam2018}.
}

% [Hypotheses / questions]
Despite the increasing use of sex selection, the critical impacts of spacing on outcomes 
for both mother and children, and the extensive use of spacing to measure son preference, 
there has so far been no systematic examination of what effect access to sex selection 
have on spacing.
I address three questions to remedy this shortcoming in the literature.

First, how did average birth intervals change over time for different groups and how do 
these changes relate to the use of sex selection?
The hypothesis is that in the absence of sex selection son preference leads to shorter 
spacing the fewer boys a family has, whereas son preference increases spacing the fewer 
boys a family has when prenatal sex selection is available, possibly to the point where
we observe a reversal of the standard spacing pattern.
Families with son preference who---for one reason or another---do not use prenatal sex 
selection likely continue to have shorter birth spacing.
Based on the prior literature on sex selection, I focus on how birth intervals have
changed for different education levels, urban or rural, and the sex composition of 
prior children.

Second, have the shapes of the survival curves changed?
Although examining average birth intervals helps us understand the overall changes in 
spacing behavior, it tells us little about how these changes came about.
For example, the motivations that lead to an increase in average spacing are different if 
the increase comes mainly from parents trying to avoid very short birth intervals or from 
continued use of sex selection to ensure a son.

Finally, how are the changes in birth intervals distributed geographically?
Given the cultural differences within India, it is unlikely that the increasing 
access to sex selection will have had the same effects throughout \citep{retherford03b}.
I, therefore, examine how birth intervals have changed across four regions.

There are four main results.
First, the intervals between births have increased for all education groups over the four 
decades.
These increases are larger the higher the parity and the higher the education 
level.
Second, those women who are most likely to use sex selection---well-educated women with no
sons---have seen the largest increases in birth intervals.
As a result, well-educated women now show a reversal of the traditional spacing 
pattern, with the longest birth intervals for women with no sons.
Third, women who are least likely to use sex selection, those with no education in
rural areas, still follow the standard pattern of short spacing when they have girls and 
little evidence of the use of sex selection.
Finally, the traditional pattern of longer average spacing with sons appear to arise from 
more very long intervals, rather than from fewer very short birth intervals; across 
education groups, there appears to be little difference in the likelihood of having a very 
short birth interval across sex compositions, especially for lower parities.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination exacerbates any bias from the 
proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
I, therefore, use a non-proportional hazard specification, which allows the shape of the 
hazard functions to vary across groups.
The use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

@@

I divide the explanatory variables into two groups.
The first group consists of characteristics that the prior literature finds affect 
spacing choice and the use of sex selection:
mother's education, sex composition of previous children, and area of residence.
The second group of variables consists of those expected to have an approximately 
proportional effect on the hazard.
These include the age of the mother at the beginning of the spell, whether the household 
owns any land, and whether the household belongs to a scheduled tribe or caste.

Women with different education levels have different hazard profiles, although the size 
and direction of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Furthermore, the use of sex selection increases with education, either because of the 
associated lower fertility or because higher income enables them to access and
afford prenatal sex determination
\citep{das_gupta97,retherford03b,jha06,Guilmoto2009a,Bongaarts2013,Portner2015b,Jayachandran2017}.%
\footnote{
It is also possible that the effect is driven by a stronger son preference, although 
higher education women have shown declining son preference \citep{bhat03,pande07}.
}
I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce the potential 
problem of including other variables as proportional.

The sex composition of previous children affects both the timing of births and the use of 
sex-selective abortions 
\citep{retherford03b,jha06,Bhalotra2008,abrevaya09,Portner2015b,Kumar2016,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
Finally, urban women have both lower fertility and, presumably, better access to
sex selection, both of which lead to higher use of sex-selective abortions than in 
rural areas \citep{retherford03b,jha06,Portner2015b}.
Area of residence is a dummy variable for the household living in an urban area.

% %
% \footnote{
% Fathers' education has two opposite predicted effects: the associated higher income
% should increase fertility and therefore lower the pressure to use sex selection, but
% the higher income also makes the use of sex selection cheaper.
% In practice, fathers' education had little effect on the hazards and the use of 
% sex-selective abortions, and I, therefore, do not include it.
% }
